Ab Initio Effective One-Electron Potential Operators. II. Applications for Exchange-Repulsion Energy in Effective Fragment Potentials

### 2019-12-13 papers/3_oep_technique.Pauli/paper-r2.tex
RWG suggestions to Paper III

\begin{abstract}
Recently, it was shown that
the effective one\hyp{}electron potentials (OEP) method
of effective elimination of the electron repulsion integrals
can be used to lower the computational cost
of evaluation of the charge transfer interaction energy
within the effective fragment potential method (EFP2)
by the factor of 20, making it no longer a bottleneck
in EFP2\hyp{}based simulations of complex systems.
Here, OEP technique 
is used to enhance computational efficiency in evaluating the
exchange\hyp{}repulsion EFP2 interaction energy
by incorporating the kinetic energy operator in the OEP
via the extended density fitting in incomplete auxiliary basis.
It is shown that the computational cost can be reduced by a factor
of 1.5 and kinetic energy integrals are no longer necessary to evaluate.
\end{abstract}

@@

\tableofcontents

@@

The exchange\hyp{}repulsion (EXR) interaction energy in the
second generation of the
effective fragment potentials (EFP2) method\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013,
   Nguyen.Pachter.Day.JCP.2014,
   Day.Jensen.Gordon.Webb.Stevens.Krauss.Garmer.Basch.Cohen.JCP.1996}
describes the non\hyp{}electrostatic repulsion
between two Hartree\hyp{}Fock\cite{Roothaan.RevModPhys.1951} 
isolated and unperturbed wavefunctions in an interacting
bi\hyp{}molecular complex.
The EXR EFP2 term was derived from the intermolecular perturbation theory
of Murrell et al.\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965,Otto.Ladik.ChemPhys.1975} 
up to second order with respect to the wavefunction overlap,
and is given by~\cite{Jensen.JCP.1996,Jensen.Gordon.MolPhys.1996,Jensen.Gordon.JCP.1998}
%
\begin{equation} \label{e:exr-efp2}
 E^{\rm Ex-Rep} \approx
 E^{\rm Ex} +
 E^{\rm Rep}(S^{-1}) + 
 E^{\rm Rep}(S^{-2}) \;.
\end{equation}
%
In the above equation,
$E^{\rm Ex}$ is the exchange energy,\cite{Jensen.JCP.1996}
%
\begin{equation} \label{e:exc-efp2}
 E^{\rm Ex} \approx 2 \sum_{i\in A}^{\rm Locc} \sum_{j\in B}^{\rm Locc} 
 \sqrt{\frac{-2\ln{\vert S_{ij}\vert}}{\pi}} \frac{S^2_{ij}}{r_{ij}} \;,
\end{equation}
%
whereas the first\hyp{} and second\hyp{}order repulsion terms
are accordingly~\cite{Jensen.Gordon.MolPhys.1996,Jensen.Gordon.JCP.1998}
%
\begin{multline} \label{e:rep-efp2.s1}
 E^{\rm Rep}(S^{-1}) \approx - 2 \sum_{i\in A}^{\rm Locc} \sum_{j\in B}^{\rm Locc} S_{ij} \\
 \times
 \left\{
 \sum_{k\in A}^{\rm Locc} F_{ik}^A S_{kj} 
+\sum_{l\in B}^{\rm Locc} F_{jl}^A S_{li} 
 -2T_{ij}
 \right\}
\end{multline}
%
and 
%
\begin{multline} \label{e:rep-efp2.s2}
 E^{\rm Rep}(S^{-2}) \approx 2 \sum_{i\in A}^{\rm Locc} \sum_{j\in B}^{\rm Locc} S_{ij}^2 
 \\ \times 
 \Bigg\{
 \sum_{x\in A}^{\rm At}  \frac{-Z_x}{r_{xb}}
+\sum_{y\in B}^{\rm At}  \frac{-Z_y}{r_{ya}} 
+\sum_{k\in A}^{\rm Locc} \frac{   2}{r_{jk}}
+\sum_{l\in B}^{\rm Locc} \frac{   2}{r_{il}}
-r_{ij}^{-1}
 \Bigg\} \;.
\end{multline}
%
The indices $i$, $j$, $k$ and $l$ label the localized occupied orbitals (LMO's, denoted by `Locc')
located at their charge centroids ${\bf r}_{i(j)}$,
whereas $x$ and $y$ label atomic nuclei (denoted by `At') with atomic numbers 
$Z_{x(y)}$ located at ${\bf r}_{x(y)}$. The relative distances are
defined by $r_{uw}=\vert {\bf r}_u-{\bf r}_w\vert$.
It is known that evaluating Eq.~\eqref{e:exr-efp2}
is remarkably efficient and gives usually very accurate
estimates of the benchmark EXR energy according to the symmetry 
adapted perturbation theory\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} (SAPT).
The computational cost of
the exchange and second\hyp{}order terms
is much less than the first\hyp{}order term,
which requires evaluaton of not only the one\hyp{}electron overlap,
but also the kinetic
energy integrals in a space of occupied molecular orbital (MO) basis.

In this work, an alternative formulation for the first\hyp{}order
term is proposed based on the effective one\hyp{}electron potential (OEP)
operator technique developed in the preceeding contribution.~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019}
It is shown that calculation of kinetic energy integrals can be avoided
by introducing OEP's.

@@

%
\begin{figure}[t]
\includegraphics[width=0.44\textwidth]{fig-1.pdf}
\caption{\label{f:fig-1} {\bf Asymptotic dependence of the exchange\hyp{}repulsion interaction energy
for selected bi\hyp{}molecular complexes.} 
%The charge transfer energy estimates are plotted for 
(a) water dimer, 
(b) water\hyp{}methanol complex, and 
(c) DMSO dimer in $S_i$ symmetry,
were one molecule has been translated
along the vector specified in the insets relative to initial geometry,
optimized at HF/6-31+G(d,p) level of theory.
The 
total interaction energy
is also shown for comparison in light grey color in this figure.
Interaction energies were obtained by using the 6-311++G(d,p)
primary basis set for (a) and (b), and 6-31++G(d,p) for (c).
In OEP calculations, EDF-1 scheme combined with small optimized auxiliary basis sets
(Supplementary Information) were used.
} 
\end{figure}
%

@@

ERI is defined by

@@

ERI's can be effectively eliminated from Eq.~\eqref{e:rep.murrell-etal.S1} 

@@

where ${\mathcal{F}}_t$ is a certain and well defined functional of ERI's.

@@

operator constructed from nuclear, Coulomb and exchange parts 
as

@@

\footnotetext[6]{Eq.~\eqref{}. Auxiliary basis sets for the EDF in $E^{\rm Rep}(S^{-1})$ 

@@

where the matrix ${\bf V}^A$ can be considered as effective fragment parameters.

@@

where 

@@

In the procedure, the required OEP matrix elements can be calculated from

@@

where ${\bf D}^X$ and ${\bf C}^X$ is the one\hyp{}particle AO density and

@@

(i) only overlap integrals, which are relatively computationally inexpensive,
need to be evaluated, instead of additional kinetic energy integrals in the EFP2
formulation that are slightly more expensive;

@@

We implemented EXR energy models from Eq.~\eqref{e:exr-efp2}
(EFP2), Eq.~\eqref{e:rep.murrell-etal.S1.oep} (OEP), as well as 
the approximate intermolecular perturbation theory with exchange of 
Murrell et~al.\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965},
%from Eq.~\eqref{e:rep.murrell-etal.S1} 
the exact EXR energy of Hayes \& Stone\cite{Hayes.Stone.MolPhys.1984}
and the density decomposition scheme 
at the Hartree\hyp{}Fock level of theory (here referred to as the DDS/HF) 
developed by Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011} 
for validation of the EDF technique.
When implemented in the dimer\hyp{}centered basis set\cite{Chalasinski.Gutowski.MolPhys.1985} (DCBS),
the DDS/HF model is
equivalent to the exact EXR energy by Hayes \& Stone and gives identical results,
which is the case in this work. Using DCBS also allows one to eliminate the basis set
superimposition error (DCBS).
To localize molecular orbitals, procedure that is necessary for the EFP2 calculations,
the Boys method\cite{Boys.RevModPhys.1960} was used. Note here that, since
Murrell et~al.'s theory is invariant with respect to unitary transformation
of molecular orbitals, the OEP\hyp{}based expression 
in Eq.~\eqref{e:rep.murrell-etal.S1.oep} is also invariant and does not require
orbital localization. 
%in which
%
%\begin{equation}\label{e:dds-exr}
% \Delta E^{\rm Ex-Rep}_{\rm DDS/HF} = 
%\end{equation}
%
%
\begin{figure*}[t]
\includegraphics[width=0.8\textwidth]{fig-2.pdf}
\caption{\label{f:fig-2} {\bf Accuracy of the OEP and EFP2 models of exchange\hyp{}repulsion energy
across various bi\hyp{}molecular systems.} 
(a) NCB31 
database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005} 
of non\hyp{}covalent interactions
and
(b) BBI subset\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
of backbone\hyp{}backbone interactions in proteins from the BioFragment Database.
For the OEP calculations, the EDF-1 scheme with the aug-cc-pVDZ-jkfit auxiliary basis set
was used.
} 
\end{figure*}
%

Three complexes: 
(H$_2$O)$_2$, 
H$_2$O--HOCH$_3$, and
(DMSO)$_2$,
were chosen as model systems to analyze the asymptotic dependence 
of EXR energy, which is shown in Figure~\ref{f:fig-1}. 
The reference (zero\hyp{}displacement) geometries
were obtained as described in Ref.\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019} 
and the structures  
along with the translation direction are depicted
in the insets in Figure~\ref{f:fig-1}.
Energy\hyp{}optimizations were performed at the HF/6-31+G(d,p) level,
as implemented in 
the {\sc Gaussian16} quantum chemistry program package.\cite{Gaussian16}
OEP and EFP2 models correctly describe the EXR energy at all separations
for the studied model systems with very similar accuracy. 
However,
contrary to our previous application of the OEP technique\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019},
where approximately 20\hyp{}fold speedups as compared to the EFP2 model were achieved
for the evaluation of the charge\hyp{}transfer energy,
the computational cost of the OEP EXR method is roughly 2 times higher than the EFP2 model
when the usual density fitting auxiliary basis sets such as aug-cc-pVDZ-jkfit or aug-cc-pVDZ-ri are used
(Table~\ref{t:tab-1}).
However, application of our small auxiliary basis sets for H$_2$O, CH$_3$OH and DMSO, 
%developed
developed via the basis set optimization~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019}
in conjunction with the basin hopping global optimization algorithm  %cite
as implemented in the SciPy Python library
(see Supplementary Information),
%to minimize the computational cost of evaluating the summation terms
%from Eq.~\eqref{e:rep.murrell-etal.S1.oep},
reduces costs appreciably and makes the OEP method on average 1.5 times faster than EFP2.
For instance, calculation of the EXR energy in the DMSO dimer by using 6-31++G(d,p) primary
basis set and 
1.2 GHz AMD EPYC\texttrademark{} 7301 16-Core Processor
requires about 4.6~ms and 3.0~ms when EFP2 and OEP model is used, respectively.
%Note that the benchmark Hayes \& Stone's calculations predict slightly larger repulsion energies

To investigate the accuracy of the OEP\hyp{}based repulsion
term across
%perform statistical error analysis for 
a variety of
interacting systems, range of
bi\hyp{}molecular complexes from the non\hyp{}covalent
interactions database NCB31 developed by the Truhlar's 
group,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
as well as the BioFragment Database subset BBI for backbone\hyp{}backbone
interactions in proteins of Sherrill group,\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
as implemented in the {\sc Psi4}
program,\cite{Psi4.JCTC.2017}
were utilized, and the
EXR energies were compared with the reference DDS/HF results (Figure~\ref{f:fig-2}).
On average, OEP and EXR tend to underestimate the EXR energy by 0.7\% and 7.6\%
in the NCB31/6-311++G(d,p) data set, and 6.0\% and 15\% in the BBI/6-31+G(d,p) data set,
respectively.
This is mostly due to the lack of BSSE, that is accounted for
only in the reference calculations with the DDS/HF model, and the effect is usually larger
when smaller primary basis sets are used.
The OEP model is of slightly better accuracy as the EFP2 model
in systems sudied in this work, with 
root mean square errors of OEP and EFP2 model being
2.0 kcal/mol and 2.6 kcal/mol for NCB31/6-311++G(d,p) data set, 
as well as
0.5 kcal/mol and 1.2 kcal/mol for BBI/6-31+G(d,p) data set, respectively.
It is an accidental fruitful cancellation of errors because
the first\hyp{}order repulsive terms have much less differences with respect to each other
(RMSE is 0.7\% only in all data sets).
Nevertheless, together with the performance data of both models
from Table~\ref{t:tab-1}, OEP model performs better than EFP2 as long as the minimal
auxiliary basis sets are used.

@@

In this work, the OEP technique of effective ERI elimination
proposed by us previously was utilized to reduce the computational cost
of evaluation of the EXR energy in the EFP2 model.
Starting from the Murrell et~al.'s model, which is a foundation of the EFP2 EXR term,
OEP\hyp{}based expression was derived which requires evaluation of only overlap
OEI's, apart from additional kinetic energy OEI's required for EFP2. 
Therefore, the following model is proposed for the
total exchange\hyp{}repulsion energy in the EFP2 model: 
(i) $E^{\rm Ex}$ and $E^{\rm Rep}(S^{-2})$ are evaluated
from the original EFP2 formulae of Jensen and Gordon, i.e.,
from Eqs.~\eqref{e:exc-efp2} and \eqref{e:rep-efp2.s2}, respectively;
(ii) $E^{\rm Rep}(S^{-1})$ term is evaluated from Eq.~\eqref{e:rep.murrell-etal.S1.oep}
and assuming a small auxiliary basis sets, optimized for each
EFP2 fragment separately.

>>>>

\begin{abstract}
In Paper~I, the effective one\hyp{}electron
potentials (OEP) method was introduced and demonstrated as an
efficient approach
to reduce the computational cost
of evaluation of the charge-transfer interaction energy
within the effective fragment potential method (EFP2)
by an average factor of 20, making it no longer a bottleneck
in EFP2\hyp{}based simulations of complex systems.
Here, the OEP technique 
is used to enhance computational efficiency in evaluating the
exchange\hyp{}repulsion EFP2 interaction energy
by redefining the first\hyp{}order repulsive term of 
Murrell~et al. [Murrell et al., Proc. R. Soc. Lond. A 284, 566 (1965)] through
the extended density fitting in incomplete auxiliary basis.
In the proposed approach the evaluation kinetic energy integrals is no
longer required and the computational cost can be reduced roughly by a factor
of 1.5 as compared to the original EFP2 formulation.
\end{abstract}

@@

%\tableofcontents

@@

The exchange\hyp{}repulsion (EXR) interaction energy in the
second generation of the
effective fragment potentials (EFP2) method\cite{Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013,
   Nguyen.Pachter.Day.JCP.2014,
   Day.Jensen.Gordon.Webb.Stevens.Krauss.Garmer.Basch.Cohen.JCP.1996}
describes the non\hyp{}electrostatic repulsion
between two isolated and unperturbed Hartree\hyp{}Fock\cite{Roothaan.RevModPhys.1951} 
wavefunctions in an interacting
bi\hyp{}molecular complex.
The EXR EFP2 term was derived from the intermolecular perturbation theory
of Murrell et al.\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965,Otto.Ladik.ChemPhys.1975} 
up to the second order with respect to the wavefunction overlap,
and is given by~\cite{Jensen.JCP.1996,Jensen.Gordon.MolPhys.1996,Jensen.Gordon.JCP.1998}
%
\begin{equation} \label{e:exr-efp2}
 E^{\rm Ex-Rep} \approx
 E^{\rm Ex} +
 E^{\rm Rep}(S^{-1}) + 
 E^{\rm Rep}(S^{-2}) \;.
\end{equation}
%
In the above equation,
$E^{\rm Ex}$ is the exchange energy,\cite{Jensen.JCP.1996}
%
\begin{equation} \label{e:exc-efp2}
 E^{\rm Ex} \approx 2 \sum_{i\in A}^{\rm Locc} \sum_{j\in B}^{\rm Locc} 
 \sqrt{\frac{-2\ln{\vert S_{ij}\vert}}{\pi}} \frac{S^2_{ij}}{r_{ij}} \;,
\end{equation}
%
whereas the first\hyp{} and second\hyp{}order repulsion terms
are accordingly~\cite{Jensen.Gordon.MolPhys.1996,Jensen.Gordon.JCP.1998}
%
\begin{multline} \label{e:rep-efp2.s1}
 E^{\rm Rep}(S^{-1}) \approx - 2 \sum_{i\in A}^{\rm Locc} \sum_{j\in B}^{\rm Locc} S_{ij} \\
 \times
 \left\{
 \sum_{k\in A}^{\rm Locc} F_{ik}^A S_{kj} 
+\sum_{l\in B}^{\rm Locc} F_{jl}^A S_{li} 
 -2T_{ij}
 \right\}
\end{multline}
%
and 
%
\begin{multline} \label{e:rep-efp2.s2}
 E^{\rm Rep}(S^{-2}) \approx 2 \sum_{i\in A}^{\rm Locc} \sum_{j\in B}^{\rm Locc} S_{ij}^2 
 \\ \times 
 \Bigg\{
 \sum_{x\in A}^{\rm At}  \frac{-Z_x}{r_{xb}}
+\sum_{y\in B}^{\rm At}  \frac{-Z_y}{r_{ya}} 
+\sum_{k\in A}^{\rm Locc} \frac{   2}{r_{jk}}
+\sum_{l\in B}^{\rm Locc} \frac{   2}{r_{il}}
-r_{ij}^{-1}
 \Bigg\} \;.
\end{multline}
%
The indices $i$, $j$, $k$ and $l$ label the localized occupied orbitals (LMO's, denoted by `Locc')
located at their charge centroids ${\bf r}_{i(j)}$,
whereas $x$ and $y$ label atomic nuclei (denoted by `At') with atomic numbers 
$Z_{x(y)}$ located at ${\bf r}_{x(y)}$. The relative distances are
defined by $r_{uw}=\vert {\bf r}_u-{\bf r}_w\vert$.
It is known that evaluating Eq.~\eqref{e:exr-efp2}
is remarkably efficient and gives usually very accurate
estimates of the reference EXR energies according to the symmetry 
adapted perturbation theory\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} (SAPT).
The computational cost to evaluate
the exchange and second\hyp{}order terms
is much smaller than that of the first\hyp{}order term,
which requires evaluation of not only the one\hyp{}electron overlap,
but also the kinetic
energy integrals in a space of occupied molecular orbital (MO) basis.

In this work, an alternative formulation for the first\hyp{}order
term is proposed based on the effective one\hyp{}electron potential (OEP)
operator technique developed in the preceding contribution.~\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019}
It is shown that using this approach calculation of kinetic energy integrals can be avoided
what leads to increased computational efficiency.

@@

%
\begin{figure}[t]
\includegraphics[width=0.44\textwidth]{fig-1.pdf}
\caption{\label{f:fig-1} {\bf Asymptotic dependence of the exchange\hyp{}repulsion interaction energy
for selected bi\hyp{}molecular complexes.} 
%The charge transfer energy estimates are plotted for 
(a) water dimer, 
(b) water\hyp{}methanol complex, and 
(c) DMSO dimer in $S_i$ symmetry,
were one molecule has been translated
along the vector specified in the insets relative to initial geometry,
optimized at HF/6-31+G(d,p) level of theory.
The 
total interaction energy
is also shown for comparison in light grey color in this figure.
Interaction energies were obtained by using the 6-311++G(d,p)
primary basis set for (a) and (b), and 6-31++G(d,p) for (c).
In OEP calculations, EDF-1 scheme combined with small optimized auxiliary basis sets
(cf. Supplementary Information) were used.
} 
\end{figure}
%

@@

ERIs are defined by

@@

ERIs can be effectively eliminated from Eq.~\eqref{e:rep.murrell-etal.S1} 

@@

where ${\mathcal{F}}_t$ is a certain and well defined functional of ERIs.

@@

operator constructed from nuclear, Coulomb and exchange
contributions as

@@

\footnotetext[6]{Same as EFP2 but with $E^{\rm Rep}(S^{-1})$ replaced by formula in Eq.~\eqref{e:rep.murrell-etal.S1.oep}. 
Auxiliary basis sets for the EDF in $E^{\rm Rep}(S^{-1})$ 

@@

where the matrix ${\bf V}^A$ can be considered as a set of effective fragment parameters.

@@

where the

@@

In this procedure, the required OEP matrix elements can be calculated from

@@

where ${\bf D}^X$ and ${\bf C}^X$ are the one\hyp{}particle AO density and

@@

(i) only overlap integrals are needed, which are relatively easy to
compute. Furthermore, there is no need to evaluate the kinetic
energy integrals, as in the original EFP2
formulation, which are computationally slightly more demanding;

@@

In our in\hyp{}house plug-in to the {\sc Psi4}
quantum chemistry
program,\cite{Psi4.JCTC.2017}
we implemented EXR energy models from Eq.~\eqref{e:exr-efp2}
(EFP2), Eq.~\eqref{e:rep.murrell-etal.S1.oep} (OEP), as well as 
the approximate intermolecular perturbation theory with exchange of 
Murrell et~al.\cite{Murrell.Randic.Williams.Longuet-Higgins.ProcRSocLondA.1965},
%from Eq.~\eqref{e:rep.murrell-etal.S1} 
the exact EXR energy of Hayes \& Stone\cite{Hayes.Stone.MolPhys.1984},
here referred to as the HS model.
The latter was
implemented in the dimer\hyp{}centered basis set\cite{Chalasinski.Gutowski.MolPhys.1985} (DCBS),
to eliminate the basis set
superposition error (DCBS) in the benchmark calculations.
%and the density decomposition scheme 
%at the Hartree\hyp{}Fock level of theory (here referred to as the DDS/HF) 
%developed by Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011} 
%for validation of the EDF technique.
%When implemented in the dimer\hyp{}centered basis set\cite{Chalasinski.Gutowski.MolPhys.1985} (DCBS),
%the DDS/HF model is
%equivalent to the exact EXR energy by Hayes \& Stone and gives identical results,
%which is the case in this work. 
%Using DCBS also allows one to eliminate the basis set
%superimposition error (DCBS).
The Boys method\cite{Boys.RevModPhys.1960} was used to localize
molecular orbitals---the same as in the original EFP2 formulation,
Note here that, since
Murrell et~al. theory is invariant with respect to unitary transformation
of molecular orbitals, the OEP\hyp{}based expression 
in Eq.~\eqref{e:rep.murrell-etal.S1.oep} is also invariant and does not require
orbital localization. 
%in which
%
%\begin{equation}\label{e:dds-exr}
% \Delta E^{\rm Ex-Rep}_{\rm DDS/HF} = 
%\end{equation}
%
%
\begin{figure*}[t]
\includegraphics[width=0.8\textwidth]{fig-2.pdf}
\caption{\label{f:fig-2} {\bf Accuracy of the OEP and EFP2 models of exchange\hyp{}repulsion energy
across various bi\hyp{}molecular systems.} 
(a) NCB31 
database,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005} 
of non\hyp{}covalent interactions
and
(b) BBI subset\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
of backbone\hyp{}backbone interactions in proteins from the BioFragment Database.
For the OEP calculations, the EDF-1 scheme with the aug-cc-pVDZ-jkfit auxiliary basis set
was used.
} 
\end{figure*}
%

Three complexes: 
(H$_2$O)$_2$, 
H$_2$O--HOCH$_3$, and
(DMSO)$_2$,
were chosen as model systems to analyze the asymptotic dependence 
of EXR energy, which is shown in Figure~\ref{f:fig-1}. 
The reference (zero\hyp{}displacement) geometries
were obtained as described in Ref.\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019} 
and the structures  
along the translation direction are depicted
in the insets in Figure~\ref{f:fig-1}.
Energy\hyp{}optimizations were performed at the HF/6-31+G(d,p) level,
as implemented in 
the {\sc Gaussian16} quantum chemistry program package.\cite{Gaussian16}
OEP and EFP2 models correctly describe the EXR energy at all separations
for the studied model systems with very similar accuracy. 
However,
contrary to our previous application of the OEP technique\cite{Blasiak.Bednarska.Choluj.Bartkowiak.JCP.2019},
where approximately 20\hyp{}fold speedups as compared to the EFP2 model were achieved
for the evaluation of the charge\hyp{}transfer energy,
the CPU timings of EXR calculations in the OEP method are comparable but roughly 2
times higher than that of the EFP2 model
when the usual density fitting auxiliary basis sets such as aug-cc-pVDZ-jkfit or aug-cc-pVDZ-ri are used
(Table~\ref{t:tab-1}).
Computational cost of $\Delta E^{\rm Rep}(S^{-1})$ in EFP2 model
is approximately $s(on^2 + o^2n) + t(on^2 + o^2n) + 2o^3$, whereas in OEP model
is $s(on^2 + o^2n) + 2ao^2$, both of which have comparable magnitudes. 
Here, $s$ and $t$ denote the relative costs of evaluation of the overlap
and kinetic-energy one electron integrals (OEIs), whereas $o$, $n$ and $a$ is
the number of LMOs, primary AOs and auxiliary AOs, respectively.
It is clear that the critical parameter is the number of auxiliary basis functions $a$,
which should be comparable with the number of occupied orbitals $o$ 
as $t$ is usually only 2--3 times larger than $s$.
To investigate this issue more thoroughly we developed small auxiliary basis sets for H$_2$O, CH$_3$OH and DMSO, 
%developed
via the basis set optimization method described in Appendix~A to Paper I
in conjunction with the basin hopping global optimization algorithm\cite{Wales.EnergyLandscapes.2003,
Wales.Scheraga.Science.1999,
Wales.Doye.JPCA.1997,Li.Scheraga.PNAS.1987}
as implemented in the SciPy Python library\cite{SciPy.2019}
(see Supplementary Information).
%to minimize the computational cost of evaluating the summation terms
%from Eq.~\eqref{e:rep.murrell-etal.S1.oep},
Applying such basis sets
reduces computational effort appreciably and makes the OEP method on average 1.5 times faster than EFP2.
For instance, calculation of the EXR energy in the DMSO dimer by using 6-31++G(d,p) primary
basis set and 
1.2~GHz AMD EPYC\texttrademark{} 7301 16-Core Processor
requires about 4.6~ms and 3.0~ms when EFP2 and OEP models are used, respectively.
%Note that the benchmark Hayes \& Stone's calculations predict slightly larger repulsion energies

To investigate the accuracy of the OEP\hyp{}based repulsion
term across
%perform statistical error analysis for 
a variety of
interacting systems, a selection of
bi\hyp{}molecular complexes from the non\hyp{}covalent
interactions database NCB31 developed by the Truhlar's 
group,\cite{Zhao.Schultz.Truhlar.JCTC.2006,
Zhao.Truhlar.JCTC.2005,Zhao.Schultz.Truhlar.JCTC.2006,Zhao.Schultz.Truhlar.JCP.2005}
as well as the BioFragment Database subset BBI for backbone\hyp{}backbone
interactions in proteins of Sherrill group,\cite{Burns.Faver.Zheng.Marshall.Smith.Vanommeslaeghe.MacKerell.Merz.Sherrill.JCP.2017} 
as implemented in the {\sc Psi4}
program,\cite{Psi4.JCTC.2017}
were utilized. The computed
EXR energies were compared with the reference HS results (Figure~\ref{f:fig-2}).
On average, OEP and EFP2 methods tend to underestimate the EXR energy by 0.7\% and 7.6\%
in the NCB31/6-311++G(d,p) data set, and 6.0\% and 15\% in the BBI/6-31+G(d,p) data set,
respectively.
This is mostly due to BSSE, that is corrected for
only in the reference calculations with the HS model, and the effect is usually larger
when smaller primary basis sets are used.
The OEP model performs slightly better when compared to the EFP2
method for systems studied in this work, with 
root-mean-square errors of OEP and EFP2 models of
2.0 kcal/mol and 2.6 kcal/mol, respectively, for NCB31/6-311++G(d,p) data set, 
or
0.5 kcal/mol and 1.2 kcal/mol for BBI/6-31+G(d,p) data set.
The slightly better performance of OEP is most likely due to
fruitful cancellation of errors, since
the mutual differences between EFP2 and OEP estimations of 
$\Delta E^{\rm Rep}(S^{-1})$ are on the average much smaller
(RMSE between EFP2 and OEP is only 0.7\% in all data sets).
%
% RWG: Aren't they increasing with the magnitude of interaction
% energy? For instance what is the difference in magnitudes for the
% last complex in NCB31 of 48kcal/mol and the last three complexes
% in BBI above 15 kcal/mol. 
% 
% From what I see the OEP is consistently closer to the reference
% values - thus I think it is a bit more than fruitful cancellation.
% Can you find one example when EFP2 is closer to the reference?
% Have you tried to compute this in other (larger) basis sets? If
% this trend would be consistent in larger basis set then it is
% pretty obvious that it is not just fruitful cancellation.
%
Nevertheless, together with the performance data of both models
from Table~\ref{t:tab-1}, OEP model consistently outperforms the
EFP2 approach in terms of accuracy and also performance, provided
that the minimal auxiliary basis set is used.

@@

In this work, the effective one-electron potential (OEP) technique,
proposed in the preceding paper for the
effective elimination of electron
repulsion integrals in ab initio calculations,
was utilized to reduce the computational cost
of evaluation of the exchange repulsion (EXR) energy in the EFP2 model.
Starting from the first-order formula for EXR in Murrell et~al. perturbation theory, 
being a foundation of the EFP2 EXR term,
OEP\hyp{}based expression was derived which requires evaluation of
only the overlap one-electron integrals, in contrast to EFP2
approach which requires also the kinetic-energy one-electron
integrals. 
The reported results indicate that a following model should be more
efficient for calculations of the 
total exchange\hyp{}repulsion energy in the EFP2 model: 
(i) $E^{\rm Ex}$ and $E^{\rm Rep}(S^{-2})$ are evaluated
from the original EFP2 formulae of Jensen and Gordon, i.e.,
from Eqs.~\eqref{e:exc-efp2} and \eqref{e:rep-efp2.s2}, respectively;
(ii) $E^{\rm Rep}(S^{-1})$ term is evaluated from Eq.~\eqref{e:rep.murrell-etal.S1.oep}
and assuming a small auxiliary basis sets, optimized for each
EFP2 fragment separately.
